---
title: "Recent Financial Cycles- Statistical Features and Parameters"
author: "Leonardo Zepeda"
date: "01/02/2021"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Abstract

By analyzing DOW and S&P indexes from 1980 to date, the author identifies 7 financial cycles and present common statistical features and parameters for each cycle. The analysis found that a cycle's length range from 764 to 1,777 days, averaging 20 basis points (bp) and 110 bp weekly and monthly returns respectively for DOW index; and 21 bp and 114 bp weekly and monthly returns  respectively for S&P index. The analysis also found that cycle's total returns averaged 1.58 and 1.59 times it initial values for DOW and S&P indexes respectively, ranging from 1.38 to 1.81 times for DOW and from 1.38 to 1.81 times for S&P. With the latest, and considering only statistical features, we could infer current cycle have between 531 to 1544 days left, with at least 11% of those days with negative returns, but with an expectation of delivering at least 1.58 times its initial values.     

## Sources of Data

The data comes from Yahoo Finance, selecting the adjusted price column. These data correct the irregularities generated by idiosyncratic events of the firms (dividends, splits, etc.) that are not market movements. Thus, the data are comparable over time. With them we create new columns: returns (daily percentage change), lagged price and lagged performance. Lagging performance is simply the performance of the previous day.


```{r, echo = FALSE}
suppressPackageStartupMessages({
    library(tidyverse)
    library(tidyquant)
    library(tibbletime)
    library(ggplot2)
    library(lubridate)
    library(matrixStats)
    library(data.table)
    library(dplyr)
    library(xts)
    library(zoo)
    library(TTR)
    library(quantmod)
    library(highcharter)
    library(timetk)
    library(cowplot)
 })
```

```{r, echo=FALSE}
DOW  <- tq_get("^DJI", get = "stock.prices", from = "1980-01-01")
SP <- tq_get("^GSPC", get = "stock.prices", from = "1980-01-01")
```

## Load and Transform Data

Our objective is to find out critical statistical features to describe a cycle: length, weekly returns, monthly returns, total returns per cycle, and other technical (statistical) features and parameters useful for risk managment.   

```{r, echo =FALSE}
df <- bind_rows("DOW" = DOW,"SP"= SP, .id = "index") %>% 
     select(date, index, adjusted) %>% 
     rename(price = adjusted)
```

## A First Visual Analysis

By observing historical levels we can notice dramatic changes against certain dramatic events. More relevant and dramatic are 1987 crisis, 9/11 in 2001 and the subprime crisis in 2009 and more recently the dramatic COVID-19 crisis in 2020. Is also noticeable that indexes behave highly correlated over time.

### Graph 1 -Levels

```{r echo=FALSE}

ggplot(df, aes(date, price, colour=index)) + 
    geom_line(data=df %>% filter(index =="DOW")) + 
    geom_line(data=df %>% filter(index =="SP"), aes(y = price*10)) + 
    scale_y_continuous(sec.axis= sec_axis(~./10, name="SP")) +
    scale_color_manual(values=c("#999999", "#CA0020", "#404040"))
```


## Leveling Perspective

Price's magnitude distract the observer from the size of the variations. By considering price levels of 2020 (around 30,000 points for DOW) the observer may percieve a reduced importance of price variations in say, 1987 (when prices of DOW were one-thenth of 2020 levels). By using returns we can focus in the magnitude of the change avoiding the distraction of increasing price levels. In order to filter only "sustained" variations, the used returns are on weekly and monthly basis

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
df <- df %>% 
     group_by(index) %>% 
     mutate(lag_price = lag(price,5),
            returnw = (price - lag_price)/ lag_price,
            lag_returnw = lag(returnw)) %>%
     na.omit()

df <- df %>% 
     group_by(index) %>% 
     mutate(lag_price = lag(price,30),
            returnm = (price - lag_price)/ lag_price,
            lag_returnm = lag(returnm)) %>%
     na.omit()

```

### Graph 2 -Weekly Returns
```{r, echo = FALSE }
df %>%
     ggplot(aes(x=date, y=returnw))+
     geom_line(aes(col=index))+
     geom_hline(yintercept=-.10, linetype="solid", 
                color = "black", size=0.2)+
     geom_hline(yintercept=.10, linetype="solid", 
                color = "black", size=0.2)+
    scale_color_manual(values=c("#999999", "#CA0020", "#404040"))

```

### Graph 3 -Monthly Returns
```{r, echo = FALSE }
df %>%
     ggplot(aes(x=date, y=returnm))+
     geom_line(aes(col=index))+
     geom_hline(yintercept=-.20, linetype="solid", 
                color = "black", size=0.2)+
     geom_hline(yintercept=.20, linetype="solid", 
                color = "black", size=0.2)+
     scale_color_manual(values=c("#999999", "#CA0020", "#404040"))
```
## Is there a trend for recovery ? 

One of the interesting aspects to analyse is to find out if there is a noticeable trend for recovery after sensitive falls. First we will analyse global paterns of recovery by ploting returns against laged returns in weekly (7 days) and monthly basis (30 days). If immediate or abrupt recovery was a pattern, we should expect a more dense cloud of points in the higher-left corner of the graph, meaning that sensitive falls corresponds to sensitive recoverys. As observed in follogiwng graph neither for DJI or S&P 500 and weekly nor monthly data sugest that it would be the case. 

##  Graph IV -Recovery Paterns
```{r, echo=FALSE }

 t1 <- 
    ggplot(df, aes(x=lag_returnw, y=returnw))+ 
    geom_point(aes(col=index), alpha = 0.4)+
    geom_abline(intercept = 0, slope = 1, lty = 3, col=3)+
    xlim(-.3,.3)+
    ylim(-.3,.3)+
    scale_color_manual(values=c("#999999", "#CA0020", "#404040"))
  


t2 <- 
   ggplot(df, aes(x=lag_returnm, y=returnm))+ 
    geom_point(aes(col=index), alpha = 0.4)+
    geom_abline(intercept = 0, slope = 1, lty = 3, col=3)+
    xlim(-.3,.3)+
    ylim(-.3,.3)+
    scale_color_manual(values=c("#999999", "#CA0020", "#404040"))

plot_grid (t1, t2, 
         labels = c("Weekly Jumps", "Monthly Jumps"),
         ncol = 2,nrow = 1)

```

To validate the latest inference, we arange data in "equally-sized volatility clusters" regardless of the time of occurence. We arange data in 100 equally sized groups. We call them Similar Volatility Groups. As a result from that arrangement, the first group will contain 1% of the smallest lagged returns, while the latest group will contain the 1% largest lagged returns. Then for each group we calculate its mean and the positive portion of each group. 

By filtering only current positive returns, we observe that the groups with the smallest returns coorespond to negative mean lagged returns and the largest returns correspond to the largest positive lagged mean returns in asyntotical line arround the 0 mean. The latest complies for both weekly and monthly lagged returns. Up to now it might be confirmed the latest inference that there is not a noticeable trend for recovery after sensitive falls.

Moreover, when ploting its density we find that there is sensitively more density of positive cases, within the 50% larger lagged returns. This may imply that the magnitude of the recovery is more often more close related with previous positive returns,  than with sensitive falls. 

We will come later again with this global result of the data.

``` {r, echo=FALSE }
N <- 100

df_cutsw <- df %>%
            group_by(index) %>%
            mutate(bin = cut(lag_returnw, 
                                  quantile(lag_returnw, 
                                           probs = seq(1 / N, 1 - 1 / N, 1 / N))) %>%
                                  as.character()) %>%
            group_by(index, bin) %>%
            summarize(pos = mean(returnw >0 ),
                      mid = mean(lag_returnw)) 


df_cutsm <- df %>%
            group_by(index) %>%
            mutate(bin = cut(lag_returnm, 
                                  quantile(lag_returnm, 
                                           probs = seq(1 / N, 1 - 1 / N, 1 / N))) %>%
                                  as.character()) %>%
            group_by(index, bin) %>%
            summarize(pos = mean(returnm>0),
                      mid = mean(lag_returnm)) 

```


## Graph V -Recovery Paterns in depth
```{r,  echo=FALSE}
tqw <- 
  ggplot(df_cutsw, aes(x=pos, y=mid))+
  geom_point(aes(col = index), alpha=0.5, size=1)+
  geom_hline(yintercept=0.00, col=2)+
  geom_vline(xintercept=0.50, col=2)+
  xlab("return group -positive portion")+
  ylab("mean of lag returns")+
  scale_color_manual(values=c("#999999", "#CA0020", "#404040"))


tqwd <- 
  ggplot() + geom_density(data=df_cutsw, aes(x=pos, group=index, color=index, ..count..), alpha=0, adjust=1)+
  theme_classic()+
  geom_vline(xintercept=0.50, col=1)+
  xlab("return group -positive portion")+
  ylab("count")+
  scale_color_manual(values=c("#999999", "#CA0020", "#404040"))

tqm <- 
  ggplot(df_cutsm, aes(x=pos, y=mid))+
  geom_point(aes(col = index), alpha=0.5, size=1)+
  geom_hline(yintercept=0.00, col=2)+
  geom_vline(xintercept=0.50, col=2)+
  xlab("return group -positive portion")+
  ylab("mean of lag returns")+
  scale_color_manual(values=c("#999999", "#CA0020", "#404040"))

tqmd <- 
  ggplot() + geom_density(data=df_cutsm, aes(x=pos, group=index, color=index,..count..), alpha=0, adjust=1)+
  theme_classic()+
  geom_vline(xintercept=0.50, col=1)+
  xlab("return group -positive portion")+
  ylab("count")+
  stat_bin()+
  scale_color_manual(values=c("#999999", "#CA0020", "#404040"))

plot_grid (tqw, tqm, tqwd, tqmd, 
         labels = c("Weekly Percentils Returns Reaction", "Monthly Percentils Returns Reaction", "Weekly Positive Returns Denisty", "Monthly Positive Returns Density"),
         ncol = 2,nrow = 2)


```  


## Analysis over Weekly Returns

### Features of cyles

There are several methods and criteria to identify cycles. 
Far from any teoretical stance and with a mere practical objective, in this analysis we will define the start of a cycle with a threshold of a 10% or bigger dayly fall, coincident with a 10% weekly and a 10% monthly fall in index prices. As arbitrary as it seems, it provides 7 coincident falls (for both inidices) from 1985 to 2020 (see Graph 2)

```{r, echo=FALSE}
dfa <- subset(df, df$returnd <-.1 & df$returnw < -.1 & df$returnm < -.1)
```

Critical dates for DJI:
1987/10/19
1998/08/31
2001/09/17
2008/10/07
2011/08/08
2015/08/25
2020/02/27

Critical dates for S&P:
1987/10/19
1998/08/31
2001/09/21
2002/07/22
2008/10/07
2009/02/23
2011/08/08
2015/08/25
2020/02/27

To identify a cycle we consider only critical dates coincident for both indexes what excludes: 2002/07/22 and 2009/02/23 from the S&P index.


### Length of Cycles

Cycle lenght is calculated by the number of days between one 10%+ (daily, weely and monthly) coincident fall and the next one in occurrence. Length cycle is representend in Graph 4. Please note that since the start date of series is arbitary, cycle one for both index are not consistent. Also please note that cycle 7 is just starting. Considering the latest we have 5 cycles to analyze. Graph 4 includes an horizontal line in 1081 days. This is the mean of cycle lenght of the 5 analyzed (complete) cycles. 

```{r, echo=FALSE}
df$cycle <- ifelse(df$date > as.Date("2020/02/27"),7, ifelse(df$date > as.Date("2015/08/25"),6 , ifelse(df$date  > as.Date("2011/08/08"), 5, ifelse(df$date  > as.Date("2008/10/07"), 4, ifelse(df$date > as.Date("2001/09/17"), 3, ifelse(df$date  > as.Date("1998/08/31"), 2,  1 ))))))
                                                                                            df <- df %>%
  group_by(index, cycle) %>%
  arrange(index,.by_group=TRUE) %>%
  mutate(days =row_number())


l <- as.data.frame(tapply(df$days,list(df$cycle,df$index), max))
 l <- l %>%  mutate (cycle = rownames(l)) 
 l <- l %>% select("cycle", "DOW", "SP")

```

### Graph 4 -Weeky Returns -Lenght of Cycles   

```{r, echo=FALSE}
l %>% 
  gather("Type", "Value", -cycle) %>%
  ggplot(aes(cycle, Value, fill= Type))+
  geom_bar(position="dodge", stat= "identity")+
  ylab("days")+
  geom_hline(yintercept=mean(l[2:6,]$DOW), linetype="solid", 
                color = "black", size=0.2)+
  scale_fill_manual(values=c("#999999", "#CA0020", "#404040"))

```

## Recovery 

### Recovery Cicles
```{r, echo =FALSE}
eDOW <- filter(df, index=="DOW" & days==1)
eDOW <- eDOW[, c(3,9)]
eDOW <- eDOW %>% select("cycle", "price")
 
eSP <- filter(df, index=="SP" & days==1)
eSP <- eSP[, c(3,9)]
eSP <- eSP %>% select("cycle", "price")

df <- df %>%
  group_by(index, cycle) %>%
  arrange(index,.by_group=TRUE) %>%
  mutate( norm= price/price[days==1][1L])
```

```{r, echo=FALSE}

p1 <- ggplot(df, aes(days, norm, colour=index)) + 
         geom_line(data=df %>% filter(index== "DOW" & cycle ==2)) + 
         geom_line(data=df %>% filter(index== "SP" & cycle ==2)) +
         geom_hline(yintercept=1.00 , linetype="solid", 
                color = "black", size=0.2)+
         scale_color_manual(values=c("#999999", "#CA0020", "#404040"))
p2 <- ggplot(df, aes(days, norm, colour=index)) + 
         geom_line(data=df %>% filter(index== "DOW" & cycle ==3)) + 
         geom_line(data=df %>% filter(index== "SP" & cycle ==3)) +
         geom_hline(yintercept=1.00 , linetype="solid", 
                color = "black", size=0.2)+
         scale_color_manual(values=c("#999999", "#CA0020", "#404040"))
p3 <- ggplot(df, aes(days, norm, colour=index)) + 
         geom_line(data=df %>% filter(index== "DOW" & cycle ==4)) + 
         geom_line(data=df %>% filter(index== "SP" & cycle ==4)) +
         geom_hline(yintercept=1.00 , linetype="solid", 
                color = "black", size=0.2)+
         scale_color_manual(values=c("#999999", "#CA0020", "#404040"))
p4 <- ggplot(df, aes(days, norm, colour=index)) + 
         geom_line(data=df %>% filter(index== "DOW" & cycle ==5)) + 
         geom_line(data=df %>% filter(index== "SP" & cycle ==5)) +
         geom_hline(yintercept=1.00 , linetype="solid", 
                color = "black", size=0.2)+
        scale_color_manual(values=c("#999999", "#CA0020", "#404040"))
p5 <- ggplot(df, aes(days, norm, colour=index)) + 
         geom_line(data=df %>% filter(index== "DOW" & cycle ==6)) + 
         geom_line(data=df %>% filter(index== "SP" & cycle ==6)) +
         geom_hline(yintercept=1.00 , linetype="solid", 
                color = "black", size=0.2)+
         scale_color_manual(values=c("#999999", "#CA0020", "#404040"))
p6 <- ggplot(df, aes(days, norm, colour=index)) + 
         geom_line(data=df %>% filter(index== "DOW" & cycle ==7)) + 
         geom_line(data=df %>% filter(index== "SP" & cycle ==7)) +
         geom_hline(yintercept=1.00 , linetype="solid", 
                color = "black", size=0.2)+
         scale_color_manual(values=c("#999999", "#CA0020", "#404040"))

plot_grid (p1, p2, p3, p4, p5, p6, 
         labels = c("1998-2001", "2001-2008", "2008-2011", "2011-2015", "2015-2020","2020-TD"),
         ncol = 2,nrow = 3)
       
```


### Weekly Returns -Descriptive statistics per cycle, 

Besides lenght of the cycle, central tendency analysis may reveal another important features. Every cycle present diferent weekly features regarding mean returns, volatitlity (standard deviation), skewness, and qurtosis. All this features are represented in the following graph and tables. We will return to this features later (when it become useful for inference and intepretation).

### Graph 5 -Weekly Returns -Distribution of returns
```{r, echo=FALSE}
ggplot(data=df, aes(x=returnw, group=cycle, color=cycle)) +
  geom_density(alpha=0.7, adjust=1) +
  theme_bw()+
  geom_vline(xintercept=0, linetype="solid", 
                color = "black", size=0.2)

 
```

```{r, echo=FALSE, results= "asis"}
s1 <- as.data.frame(tapply(df$returnw,list(df$cycle,df$index), mean)*10000)
s2 <- as.data.frame(tapply(df$returnw,list(df$cycle,df$index), sd))
s3 <- as.data.frame(tapply(df$returnw,list(df$cycle,df$index), skewness))
s4 <- as.data.frame(tapply(df$returnw,list(df$cycle,df$index), kurtosis))
DOWs <- cbind(l$DOW, s1$DOW, s2$DOW, s3$DOW, s4$DOW)
rownames(DOWs) <- rownames(l)
colnames(DOWs) <- c("days", "mean(bp)", "sd", "skewness", "kurosis")
SPs <- cbind(l$SP, s1$SP, s2$SP, s3$SP, s4$SP)
rownames(SPs) <- rownames(l)
colnames(SPs) <- c("days", "mean(bp)", "sd", "skewness", "kurosis")

DOWs <- as.data.frame(DOWs)
SPs <- as.data.frame(SPs)

DOWs

SPs

```

### Monthly Returns 


### Monthly Returns -Descriptive statistics per cycle, 

Besides lenght of the cycle, central tendency analysis may reveal another important features. Every cycle present diferent monthly features regarding mean returns, its volatitlity (standard deviation), its skewness, and qurtosis. All this features are represented in the following graph and tables. We will return to this features later (when it become useful for inference and intepretation).

### Graph 7 -Distribution of returns
```{r, echo=FALSE}
ggplot() + geom_density(data=df, aes(x=returnm, group=cycle, color=cycle), alpha=0.8, adjust=3)+
  theme_bw()+
  geom_vline(xintercept=0, linetype="solid", 
                color = "black", size=0.2)
```

```{r, echo=FALSE, results= "asis"}
s1m <- as.data.frame(tapply(df$returnm,list(df$cycle,df$index), mean)*10000)
s2m <- as.data.frame(tapply(df$returnm,list(df$cycle,df$index), sd))
s3m <- as.data.frame(tapply(df$returnm,list(df$cycle,df$index), skewness))
s4m <- as.data.frame(tapply(df$returnm,list(df$cycle,df$index), kurtosis))
DOWsm <- cbind(l$DOW, s1m$DOW, s2m$DOW, s3m$DOW, s4m$DOW)
rownames(DOWsm) <- rownames(l)
colnames(DOWsm) <- c("days", "mean(bp)", "sd", "skewness", "kurosis")
SPsm <- cbind(l$SP, s1m$SP, s2m$SP, s3m$SP, s4m$SP)
rownames(SPsm) <- rownames(l)
colnames(SPsm) <- c("days", "mean(bp)", "sd", "skewness", "kurosis")

DOWsm <- as.data.frame(DOWsm)
SPsm <- as.data.frame(SPsm)

DOWsm
SPsm

```



```{r, echo=FALSE}
neg <- as.data.frame(tapply(df$norm<1 ,list(df$cycle,df$index), count))
neg <- neg %>% mutate( cycle= rownames(neg))
negP <- cbind(as.data.frame(neg$DOW*100/l$DOW), as.data.frame(neg$SP*100/l$SP))
names(negP) <- c("DOW","SP")
negP <- negP %>% mutate( cycle= rownames(neg))

neg = gather(neg, key = var, value = value, DOW, SP)
ggplot(neg, aes(x = cycle, y = value, fill = var)) +
  geom_bar(stat = 'identity', position = 'dodge')+
  geom_hline(yintercept=mean(neg[neg$var=="DOW",3]), linetype="solid", 
                color ="#404040", size=0.2)+
  geom_hline(yintercept=mean(neg[neg$var=="SP",3]), linetype="solid", 
                color = "#CA0020", size=0.2)+
  labs(title = "Days with negative returns per cycle")+
  scale_fill_manual(values=c("#999999", "#CA0020", "#404040"))
 
  

negP = gather(negP, key = var, value = value, DOW, SP)
ggplot(negP, aes(x = cycle, y = value, fill = var)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  geom_hline(yintercept=mean(negP[neg$var=="DOW",3]), linetype="solid", 
                color =  "#404040", size=0.2)+
  geom_hline(yintercept=mean(negP[negP$var=="SP",3]), linetype="solid", 
                color = "#CA0020", size=0.2)+
  labs( title= "% of days with negative returns per cycle")+
  scale_fill_manual(values=c("#999999", "#CA0020", "#404040"))

```

Achieve <- tapply(df$norm, list(df$index, df$cycle), FUN=max)
Achieve <- Achieve[, c(2:6)]





